---
title: "STAT574E Group 3 Final Project Appendix"
author: "Alex Salce, Raymond Owino, Matthew Wallace"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    theme: cosmo
  pdf_document: default
header-includes:
- \renewcommand*\familydefault{\sfdefault} %% this picks a sans serif font
- \usepackage[T1]{fontenc}
---

```{r warning=FALSE, message=FALSE, echo=FALSE}

library(dplyr)
library(tigris)
library(sf)
library(ggplot2)
library(mapview)
library(tidyverse)
library(scales)
library(knitr)

```

```{r datasetload, warning=FALSE, message=FALSE, echo=FALSE}

# load datasets
library(here)
set_here(path='../../')

load(here('data/wfigs_az_sf_EPSG32612.RData'))
load(here('data/az_rd_4WD_EPSG32612.RData'))
load(here('data/az_rd_primary_EPSG32612.RData'))
load(here('data/az_rd_secondary_EPSG32612.RData'))


# Load TIFs if needed

# library(raster)
# library(sp)
# 
# az_slope_data <- raster(here('data/AZ_BAEA_DEM_snapped_rsmpld_NAD83_12N.tif'))
# 
# coords <- data.frame(x = c(500000, 600000), y = c(3800000, 3900000))
# values <- extract(az_slope_data, coords)

#NOTE: all sf object data uses CRS ESPG:32612
# https://epsg.io/32612


```

```{R azstatesf}

arizona_sf <- states() %>% filter_state("arizona")
arizona_sf <- st_transform(arizona_sf, crs = "EPSG:32612")

az_counties_sf <- counties(state = "AZ", cb = TRUE)
az_counties_sf <- st_transform(az_counties_sf, crs = "EPSG:32612")

```


### Create Wildfire Dataframes

```{r finaldataframes}

az_fires_rx <- wfigs_az_sf %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf %>% filter(IncidentTypeCategory=="WF")

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")


```


## Histograms

### Nearest Roads

This histogram uses input `sf` object to create `histdf`, which is used to generate a histogram of the number of closest roads, either primary/secondary roads or 4wd roads, to each wildfire incident origin. The $x$-axis indicates the distance in meters of that bin to wildfire points.

As a generalization, we will think of Primary & Secondary roads as essentially the same class of roads, and we know that many humans use these roads every day. No matter the stretch of road, there is a good chance of there being some human settlement of some kind alongside these roads within close proximity.

The "4WD" roads are technically a fairly "incomplete" dataset, but they do give a sense of areas humans can spend their time in remote areas. In general, though, if a wildfire incident is closer to a 4WD road we can think of it as being a "more remote" fire, i.e. there is a good chance it's fairly removed from any prominent human settlement.

For reference, the minimum distances as well as the closest road type indicators were generated using the below code.

```{r nearestrdDataGen, eval=FALSE}

wfigs_az_sf$distance_rd_primary <-
  st_distance(wfigs_az_sf, az_rd_primary) %>% apply(1, min)

wfigs_az_sf$distance_rd_secondary <-
  st_distance(wfigs_az_sf, az_rd_secondary) %>% apply(1, min)

wfigs_az_sf$distance_rd_4wd <-
  st_distance(wfigs_az_sf, az_rd_4wd) %>% apply(1, min)


wfigs_az_sf <- wfigs_az_sf %>%
  mutate(distance_rd_min_prisec = pmin(distance_rd_primary, 
                                    distance_rd_secondary))

wfigs_az_sf$distance_rd_min_isprisec <- as.integer(wfigs_az_sf$distance_rd_min_all ==
                                                     wfigs_az_sf$distance_rd_min_prisec) 

```

#### All wildfire data

We will first look at the "Nearest Roads" histograms for all wildfire data (no filtering for the size of the fire.)

```{r disthistsmindistprimsec4wdNothresh, echo=FALSE}


bins <- 30

# human caused

histdf1 <- az_fires_wf_hum

# naturally caused

histdf2 <- az_fires_wf_nat

# undetermined / unknown cause

histdf3 <- az_fires_wf_un


x_rng <- range(c(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
                 histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0],
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0]))
y_max <- max(hist(histdf1$distance_rd_min_all, plot = FALSE)$counts, 
             hist(histdf2$distance_rd_min_all, plot = FALSE)$counts)


#overlay hist of dist from roads

# one plot
# par(mfrow=c(1, 2))

hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to Human caused fires", xlab = "Distance (m)")
hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to naturally caused fires", xlab = "Distance (m)")
hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))



df_names <- c("Human Caused Fires", "Naturally Caused Fires", "Unknown Caused Fires")
row_counts <- c(nrow(az_fires_wf_hum), nrow(az_fires_wf_nat), nrow(az_fires_wf_un))

# Create a table
result_table <- data.frame(WildFireType = df_names, Counts = row_counts)

kable(result_table, label = "Wildfire Counts")

rm(histdf1, histdf2, histdf3, x_rng, y_max, bins, df_names, row_counts)

```

There is a very high concentration of wildfires cause by humans very close to primary and secondary roads, and natural fires generally seem to be more remote (in general closer to more 4WD roads.)


#### Thresholded for Wildfire Acreage

We define the threshold for a fire to be "large" as `IncidentSize` $\geq 1000$ acres. We threshold the data accordingly and replot our histograms.

```{r thresholding_1}

####################################
# RUN BELOW FOR FIRE SIZE THRESHOLD#
####################################


# fire size threshold
# FIRE_SIZE_CLASS = Code for fire size based on the number of acres within the 
# final fire perimeter (A=greater than 0 but less than or equal to 0.25 acres, 
# B=0.26-9.9 acres, C=10.0-99.9 acres, D=100-299 acres, E=300 to 999 acres, 
# F=1000 to 4999 acres, and G=5000+ acres).

# class F and G fires
wf_size_threshold <- 1000

# max prescribed burn size
# wf_size_threshold <- max(az_fires_rx$IncidentSize)

wfigs_az_sf$size_threshold <- as.integer(wfigs_az_sf$IncidentSize >= wf_size_threshold)

wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(size_threshold==1)
# wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(IncidentSize >= wf_size_threshold)


az_fires_rx <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="WF")


#human caused wildfires

table(wfigs_az_sf_thresh$FireCause)

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```

```{r disthistsmindistprimsec4wdThresh, echo=FALSE}

bins <- 30

# human caused

histdf1 <- az_fires_wf_hum

# naturally caused

histdf2 <- az_fires_wf_nat

# undetermined / unknown cause

histdf3 <- az_fires_wf_un


x_rng <- range(c(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
                 histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0],
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0]))
y_max <- max(hist(histdf1$distance_rd_min_all, plot = FALSE)$counts, 
             hist(histdf2$distance_rd_min_all, plot = FALSE)$counts)


#overlay hist of dist from roads

# one plot
# par(mfrow=c(1, 2))

hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to Human caused fires", xlab = "Distance (m)")
hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to naturally caused fires", xlab = "Distance (m)")
hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

df_names <- c("Human Caused Fires", "Naturally Caused Fires", "Unknown Caused Fires")
row_counts <- c(nrow(az_fires_wf_hum), nrow(az_fires_wf_nat), nrow(az_fires_wf_un))

# Create a table
result_table <- data.frame(WildFireType = df_names, Counts = row_counts)

kable(result_table, label = "Wildfire Counts")

rm(histdf1, histdf2, histdf3, x_rng, y_max, bins, df_names, row_counts)

```



## Models

### Spatial Linear Model - Raymond

```{r warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}

library(spmodel)

```

```{r splm, warning=FALSE, cache=TRUE, eval=FALSE}

az_fires_wf_lm <- az_fires_wf_nat

az_wf_splm_esv <- esv(IncidentSize ~ distance_rd_min_prisec + tmax + tmin + prcp + mean_slope + mean_forest + pop.density, data = az_fires_wf_lm)

plot(az_wf_splm_esv)

az_wf_splm <- splm(IncidentSize ~ distance_rd_min_prisec + tmax + tmin + prcp + mean_slope + mean_forest + I(log(pop.density)), data = az_fires_wf_lm, spcov_type = "matern", estmethod = "sv-wls")

summary(az_wf_splm)

```


### Point Process Spatial Intensity Model - Matthew

Checking K, F, G functions to determine if wildfires are CSR or not.


### Point Process Binary GLM - Alex

#### Threshold for large wildfires

```{r thresholding_2}

####################################
# RUN BELOW FOR FIRE SIZE THRESHOLD#
####################################


# fire size threshold
# FIRE_SIZE_CLASS = Code for fire size based on the number of acres within the 
# final fire perimeter (A=greater than 0 but less than or equal to 0.25 acres, 
# B=0.26-9.9 acres, C=10.0-99.9 acres, D=100-299 acres, E=300 to 999 acres, 
# F=1000 to 4999 acres, and G=5000+ acres).

# class F and G fires
wf_size_threshold <- 1000

# max prescribed burn size
# wf_size_threshold <- max(az_fires_rx$IncidentSize)

wfigs_az_sf$size_threshold <- as.integer(wfigs_az_sf$IncidentSize >= wf_size_threshold)

wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(size_threshold==1)
# wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(IncidentSize >= wf_size_threshold)


az_fires_rx <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="WF")


#human caused wildfires

table(wfigs_az_sf_thresh$FireCause)

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```



#### Counties

```{r cococounty}

# Extract Coconino County
coconino_sf <- az_counties_sf %>% 
  filter(NAME == "Coconino")

#filter large wildfire data to only Coconino County
az_fires_wf <- st_intersection(az_fires_wf[az_fires_wf$FireCause=="Human" | az_fires_wf$FireCause == "Natural",], coconino_sf)

```

#### Model

```{r sglmlibrary, message=FALSE, warning=FALSE}

library(spatstat)
library(spmodel)
library(lubridate)
library(tidycensus)

```

```{r refreshfilter, echo=FALSE}

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```

```{r builddataspglmppp, cache=TRUE}

# set seed to 219 for report!
set.seed(219)

# set seed to 20 for predictions!
# set.seed(20)

# pick df of interest

az_fires_wf_spglm <- az_fires_wf_nat


az_fires_wf_spglm_ppp <- as.ppp(az_fires_wf_spglm)

background <- rpoispp((az_fires_wf_spglm_ppp$n) / area(as.owin(az_fires_wf_spglm_ppp)),
                      win = as.owin(coconino_sf))

df <- data.frame(x = background$x, y = background$y)

background_sf <- st_as_sf(df, coords = c("x", "y"), crs = "EPSG:32612")

```

```{r buildbkgcovs, warning=FALSE, message=FALSE}

## Build covariate data for background points

background_sf$distance_rd_primary <-
  st_distance(background_sf, az_rd_primary) %>% apply(1, min)

background_sf$distance_rd_secondary <-
  st_distance(background_sf, az_rd_secondary) %>% apply(1, min)

background_sf$distance_rd_4wd <-
  st_distance(background_sf, az_rd_4wd) %>% apply(1, min)


# dist_road covariates

background_sf <- background_sf %>%
  mutate(distance_rd_min_all = pmin(distance_rd_primary, 
                                    distance_rd_secondary, 
                                    distance_rd_4wd))

background_sf <- background_sf %>%
  mutate(distance_rd_min_prisec = pmin(distance_rd_primary, 
                                    distance_rd_secondary))

background_sf$distance_rd_min_isprisec <- as.integer(
  background_sf$distance_rd_min_all ==
    background_sf$distance_rd_min_prisec) 

# generate random dates

start_date <- ymd_hms("2014-01-01 00:00:00")
end_date <- now()
n <- nrow(background_sf)

random_dates <- sample(seq(start_date, end_date, by = "min"), n, replace = TRUE)
random_dates <- as.POSIXlt(random_dates)

background_sf$FireDiscoveryDateTime <- random_dates


## Population density data

population_data <- get_decennial(
  geography = "tract",
  variables = c(population="P001001"),
  state = "AZ",
  year = 2010,
  geometry = TRUE
)
population_data$variable = population_data$value/st_area(population_data$geometry)
names(population_data) <- c("GEOID","NAME","pop.density","pop.","geometry")

population_data <- st_transform(population_data, crs = "EPSG:32612")

background_sf <- st_join(background_sf, population_data, left = TRUE)


```


```{r gencsvforRaymond, eval=FALSE}

coords <- st_coordinates(background_sf)
df <- st_drop_geometry(background_sf)
df$x <- coords[, 1]
df$y <- coords[, 2]
write.csv(df, file = 'predictions_sf.csv', row.names = FALSE)

rm(df, coords)


```

```{r combine bkgwfcovariates}

df1 <- st_drop_geometry(az_fires_wf_spglm)
df2 <- st_drop_geometry(background_sf)

common_cols <- intersect(names(df1), names(df2))

df1 <- df1[, c(common_cols)]
df2 <- df2[, c(common_cols)]

# NEED TO FIX DATA TYPES IN NEW COLUMNS TO BE ABLE TO BIND
Covariates <- rbind(df1, df2)

rm(df1,df2,df)

```


```{r}

all_points <- superimpose(unmark(az_fires_wf_spglm_ppp), background)

Wildfires <- c(rep(1, az_fires_wf_spglm_ppp$n), rep(0, background$n))

data <- cbind(Wildfires, Covariates)

data_sf <- st_as_sf(cbind(data, as.data.frame(all_points)[, c('x', 'y')]),
coords = c('x', 'y'))

```

```{r}
# model formula


spglm_formula <- Wildfires ~ distance_rd_min_prisec + distance_rd_4wd + distance_rd_min_isprisec

az_wf_spglm <- spglm(spglm_formula, data = data_sf, family = binomial, spcov_type = "matern")


summary(az_wf_spglm)

```

